Title Goes Here \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2021-04-06 analysis/paper/shceof.tex
Renderiza

We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 70°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum. We chose this latitude band because it is wide enough to capture most of the relevant anomalies.

@@

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability. This motivates the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign. For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}. Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S. This also nearly minimises the correlation with the imaginary part.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
\end{figure}

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2}. The first mode (CEOF1) explains 82\% of the variance, while the second mode (CEOF1) explains a smaller fraction (7\%). In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.

@@

\begin{figure}
\centering
\includegraphics{../figures/eof1-regr-gh-1.pdf}
\caption{\label{fig:eof1-regr-gh}Regression coefficients of the real and imaginary part of the first complex EOF on SON geopotential height for the 1979 -- 2019 period. These coefficients come from multiple linear regression involving the real and imaginary part of both EFO2.}
\end{figure}

Figure~\ref{fig:eof1-regr-gh} shows regression patterns between geopotential height anomalies and cEOF1. At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

@@

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa (Fig.~\ref{fig:eof-naive}), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

\begin{figure}
\centering
\includegraphics{../figures/eof2-regr-gh-1.pdf}
\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second EOF.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.}
\end{figure}

@@

The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS. Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

@@

Figure~\ref{fig:o3-regr} shows regression maps of EOF1 with fields of Total Ozone Column (TOC). It shows zonal wave 1 pattens in TOC associated with both phases of EOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}. Thus, the Real EOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary EOF1 is shifted by 90º.

\begin{figure}
\centering
\includegraphics{../figures/wave1-o3-1.pdf}
\caption{\label{fig:wave1-o3}Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 70°S and 45°S for the period 1979 -- 2019.}
\end{figure}

Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS \textcolor{red}{(Habría que justifica mejor estas elecciones de bandas latitudinales a paritr del gráfico anterior)} and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93). As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

@@

\begin{figure*}
\includegraphics{../figures/psa-eof2-1} \caption{Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}. For the period 1979 -- 2019.}\label{fig:psa-eof2}
\end{figure*}

The relationship between PSAs and the modes described from the cEOFs is also studied. Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2. As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary EOF2 (Figure~\ref{fig:psa-eof2}b.1), and between PSA2 and Real EOF2 (Figure~\ref{fig:psa-eof2}a.2). Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure~\ref{fig:psa-eof2} panels a.1 and b.2). So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2. It is also the same that maximises this clean association between EFO2 parts and PSA modes.

\textcolor{red}{(Lo del histograma habría que explicarlo con un pocoo más detalle porque no es tan trivial. Primero decir para qué se hace, qué se hace exactamente y después recien decir cuál es el resultado y qué información aporta)}

The reason that the EOF analysis arrives to this optimal separation is that the imaginary cEOF is the most common phase (Figure~\ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram showing phase distribution of EOF2 for the period 1979 -- 2019. \textcolor{red}{(También ampliar)}}
\end{figure}

By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence of XXX. This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~@ref:(phase-histogram) with their Figure 6). \textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)} The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation. As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.

@@

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}. Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
\end{figure}

@@

\textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
The split between A-SAM and S-SAM gives more insight into the nature of the relationship. The relationship between the SAM and ¿the imaginary EOF1? (Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.

The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).

\begin{figure}
\centering
\includegraphics{../figures/eof-filtered-1.pdf}
\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary units.}
\end{figure}

@@

As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed. The resulting pattern is an nearly zonally symmetrical annular mode. This mode is highly correlated with SAM (0.75 (CI: 0.58 -- 0.86)), but only with the symmetrical part (0.88 (CI: 0.79 -- 0.94)). Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
\end{figure}

This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA. Figure~\ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes. If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2. Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric EOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.

@@

\begin{figure*}
\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b). Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}\label{fig:sst-psi-2}
\end{figure*}

Figure~\ref{fig:sst-psi-2} shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa. The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2). This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.86). Streamfunction anomalies show a coherent picture. The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2). This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.

Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location. However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. The correlation between the Real EOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}. This lack of tropical signal suggests different natures of the different phases of the EOF2 wave train.

\begin{figure}
\centering
\includegraphics{../figures/enso-phase-1.pdf}
\caption{\label{fig:enso-phase}Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019. Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile. Black line is the fit ONI \textasciitilde{} cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.}
\end{figure}

To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of EOF2 is above the median. In years with positive ONI, the phase of the EOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa. In years with near neutral ENSO, the phase of the EOF2 is much more variable. The black line in Figure~\ref{fig:enso-phase} is an harmonical fit of the relationship between ONI and cEOF2 phase. The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.

Furthemore, Figure~\ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest EOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2). Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.

It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs. In the former case, the wave train has little phase preference. However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase. This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.

\begin{figure*}
\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for EOF1.}\label{fig:sst-psi-1}
\end{figure*}

Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for EOF1. There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1. Consistently, streamfunction anomalies do not show any tropical influence. Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.

@@

\begin{figure}
\centering
\includegraphics{../figures/pp-america-1.pdf}
\caption{\label{fig:pp-america}Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) EOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2018 period. Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}
\end{figure}

>>>>

We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum. We chose this latitude band because it is wide enough to capture most of the relevant anomalies.

@@

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability. This motivates the decision of performing complex EOF jointly between levels. The computation of the cEOFs was carried out using data from both levels at the same time, therefore, each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign. For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}. Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S. This also nearly minimises the correlation with the imaginary part.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Time series of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
\end{figure}

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2}. The first mode (CEOF1) explains 82\% of the variance, while the second mode (CEOF1) explains a smaller fraction (7\%). In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.

@@

\begin{figure}
\centering
\includegraphics{../figures/eof1-regr-gh-1.pdf}
\caption{\label{fig:eof1-regr-gh}Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height for the 1979 -- 2019 period. These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.}
\end{figure}

Figure~\ref{fig:eof1-regr-gh} shows regression patterns between geopotential height anomalies and cEOF1. At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

@@

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa (Fig.~\ref{fig:eof-naive}), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

\begin{figure}
\centering
\includegraphics{../figures/eof2-regr-gh-1.pdf}
\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second cEOF.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 75°S and 45°S for the period 1979 -- 2019.}
\end{figure}

@@

The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S. Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

@@

Figure~\ref{fig:o3-regr} shows regression maps of cEOF1 with fields of Total Ozone Column (TOC). It shows zonal wave 1 pattens in TOC associated with both phases of cEOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}. Thus, the Real cEOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.

\begin{figure}
\centering
\includegraphics{../figures/wave1-o3-1.pdf}
\caption{\label{fig:wave1-o3}Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 75°S and 45°S for the period 1979 -- 2019.}
\end{figure}

Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between 75°S and 45°S; as shown in Figure~\ref{fig:o3-regr}, this is a wide latitude band that captures most of the area with significant regression coefficients. The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93). As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).

@@

\begin{figure*}
\includegraphics{../figures/psa-eof2-1} \caption{Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}. For the period 1979 -- 2019.}\label{fig:psa-eof2}
\end{figure*}

The relationship between PSAs and the modes described from the cEOFs is also studied. Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2. As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2 (Figure~\ref{fig:psa-eof2}b.1), and between PSA2 and Real cEOF2 (Figure~\ref{fig:psa-eof2}a.2). Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2 (Figure~\ref{fig:psa-eof2} panels a.1 and b.2). So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

This particular rotation of cEOF2 is the one which maximises the relationship between ENSO and Imaginary cEOF2. It is also the same that maximises this clean association between cEOF2 parts and PSA modes.

The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal. Figure~\ref{fig:phase-histogram} is an histogram that counts the number of SON years in which the cEOF2 was near a particular phase with the observations marked as rugs on the horizontal axis. For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase. About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the period 1979 -- 2019. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.}
\end{figure}

This phase-locking behaviour indicates that the Imaginary phase both is the direction that has the maximum relationship with ENSO and is also the most common phase. Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional cEOF analysis will tend to separate the two.

This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)}

The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation. As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.

@@

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}. Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
\end{figure}

@@

\textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
The split between A-SAM and S-SAM gives more insight into the nature of the relationship. The relationship between the SAM and ¿the imaginary cEOF1? (Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.

The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).

\begin{figure}
\centering
\includegraphics{../figures/eof-filtered-1.pdf}
\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º. Arbitrary units.}
\end{figure}

@@

As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed. The resulting pattern is an nearly zonally symmetrical annular mode. This mode is highly correlated with SAM (0.75 (CI: 0.58 -- 0.86)), but only with the symmetrical part (0.88 (CI: 0.79 -- 0.94)). Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the cEOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
\end{figure}

This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA. Figure~\ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes. If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2. Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.

@@

\begin{figure*}
\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b). Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}\label{fig:sst-psi-2}
\end{figure*}

Figure~\ref{fig:sst-psi-2} shows regression maps between cEOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa. The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2). This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.86). Streamfunction anomalies show a coherent picture. The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2). This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.

Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location. However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. The correlation between the Real cEOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}. This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.

\begin{figure}
\centering
\includegraphics{../figures/enso-phase-1.pdf}
\caption{\label{fig:enso-phase}Relationship between ENSO and phase of cEOF2 for the period 1979 -- 2019. Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile. Black line is the fit ONI \textasciitilde{} cos(phase) + sin(phase) computed by OLS weighted by the magnitude of cEOF2.}
\end{figure}

To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the cEOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median. In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa. In years with near neutral ENSO, the phase of the cEOF2 is much more variable. The black line in Figure~\ref{fig:enso-phase} is an harmonical fit of the relationship between ONI and cEOF2 phase. The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.

Furthemore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2). Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.

It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs. In the former case, the wave train has little phase preference. However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase. This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.

\begin{figure*}
\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for cEOF1.}\label{fig:sst-psi-1}
\end{figure*}

Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for cEOF1. There is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1. Consistently, streamfunction anomalies do not show any tropical influence. Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.

@@

\begin{figure}
\centering
\includegraphics{../figures/pp-america-1.pdf}
\caption{\label{fig:pp-america}Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2018 period. Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}
\end{figure}
